Plasticity and adaptation to high light intensity amplify the advantage of amphistomatous leaves

### 2025-08-14 ms/ms.qmd
proofread supporting materials

We compared \aax{} among `r n_acc` ecologically diverse populations of wild tomato, including representatives of all described species of *Solanum* sect. *Lycopersicon* and sect. *Lycopersicoides* [@peralta_taxonomy_2008] and the cultivated tomato *S. lycopersicum* var. *lycopersicum* (@tbl-populations). These populations were selected in a previous study because they encompass the breadth of climatic variation in the wild tomato clade [@pease_phylogenomics_2016]. Due to constraints on growth space and time, we spread out measurements over `r n_weeks` weeks from `r start_licor_date` to `r end_licor_date`. Replicates within population were evenly spread out over this period to prevent confounding of temporal variation in growth conditions with accession.

@@

We recorded $\mathrm{PPFD}$ using full spectrum quantum sensors (SQ-500-SS, Apogee Instruments, Logan, Utah, USA); we recorded temperature, RH, and [CO$_2$] using an EE894 sensor (E+E Elektronik, Engerwitzdorf, Austria) protected by a radiation shield. All environmental measurements were taken every 10 minutes from the middle of plants racks at approximately the same height as the leaves we measured. We measured leaf temperature of focal leaves prior to measurement using an infrared radiometer (SI-111-SS, Apogee Instruments, Logan, Utah, USA).

@@

Seeds provided by the Tomato Genetics Resource Center germinated on moist paper in plastic boxes after soaking for 30-60 minutes in a 50% (volume per volume) solution of household bleach and water, followed by a thorough rinse. We transferred seedlings to cell-pack flats containing Pro-Mix BX potting mix (Premier Tech, Rivière-du-Loup, Quebec, Canada) once cotyledons fully emerged, typically within 1-2 weeks of sowing. We grew seeds and seedlings for both sun and shade treatments under the same environmental conditions (12:12 h, `r germ_temp_day`:$\qty{`r germ_temp_night`}{\degreeCelsius}$, `r germ_rh_day`:`r germ_rh_night` RH day:night cycle). LED light provided $\mathrm{PPFD} = \qty{`r germ_ppfd`}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$ (Fluence RAZRx, Austin, Texas, USA).

@@

#### Light-saturated photosynthetic rate

In `r percent_aa_and_aq`\% of plants, we measured the maximum photosynthetic rate ($A_\mathrm{max}$) on the same leaflets as \agcurve{} curves. However, when a leaflet was damaged during \agcurve{} curves, we used the next closest leaflet for $A_\mathrm{max}$. Leaves acclimated to high light-intensity (\ppfdequals{2000}), ambient CO$_2$ (\caequals{415}), \rhequals{50}, and \tleafequals{25}. After $A$ and \gsw{} stabilized, we measured $A$ at $\qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$.

@@

1. Estimate \agcurve{} curve parameters for each leaf
2. Estimate \aax{} for both measurement light intensities using \agcurve{} curve parameters
3. Estimate the effects of light intensity, light treatment, and population on \aax{} (acclimation and plasticity hypotheses)
4. Estimate the effects of native light habitat on population-level \aax{} (constitutive hypothesis)

All models were fit using a Bayesian model with HMC sampling in the probabilistic programming language *Stan* [@stan_development_team_stan_2025] using the *R* package **brms** version `r fit_aa1$version$brms` [@burkner_brms_2017]. We used *CmdStan* version `r fit_aa1$version$cmdstan` and **cmdstanr** version `r fit_aa1$version$cmdstanr` [@gabry_cmdstanr_2025] to interface with *R* version `r getRversion()` [@r_core_team_r:_2025]. We sampled the posterior distribution from a single chain with a minimum of `r aa_args$n_iter_init / 2` iterations after `r aa_args$n_iter_init / 2` warmup iterations. If necessary, we refit models with more iterations to decrease the Gelman-Rubin convergence statistic ($\hat{R}$) [@gelman_inference_1992] to less than `r aa_args[["max_rhat"]]` and the bulk effective sample size (ESS) to greater than `r aa_args[["min_ess"]]` for all model parameters, and there were fewer than `r aa_args[["max_divergent"]]` divergent transitions.

@@

We fit Bayesian mixed effects models with phylogenetically structured random effects to test predictions of competing hypotheses about why amphistomy advantage (\aax) might be greater for leaves in sunny, open habitats. We fit all models combinations of measurement light intensity, growth light intensity, and their interaction as both fixed effects and random effects among populations. In other words, we test whether there were main effects of light intensity treatments and/or whether populations varied in their response to light intensity treatments. In all models, we included population as a phylogenetically structured random effect. We accounted for uncertainty in $\widehat{\mathrm{AA}}$ using the standard deviation of the posterior distribution of each $\widehat{\mathrm{AA}}$ estimate (@sec-aa-est). We crossed fixed and random effect structures with fixed effects of measurement light intensity and/or growth light intensity on the distributional parameter, $\sigma$, the residual variance in \aax{}. We used a robust regression approach by assuming $t$-distributed residuals, which makes estimates less susceptible to extreme values. In total, we fit `r n_model1` models, each with a different combination of fixed, random, and distributional effects. We used the RAxML whole-transcriptome concatenated phylogeny based on 2,745 100-kb genomic windows from @pease_phylogenomics_2016. Two of our populations were not in this tree. We used accession LA1044 in place of LA3909, two populations of *S. galapagnse*; LA0750 was added as sister to LA0716, two closely related populations of *S. pennellii*. The node separating LA0750 from LA0716 was placed half-way between the next deepest node. We used the leave-one-out cross-validation information criterion (LOOIC) to compare the fit of models (\autoref{tbl-aa_loo1}) using the *R* package **loo** version `r packageVersion("loo")` [@vehtari_practical_2017] to calculate LOOIC values. We selected the model with the lowest LOOIC value (@tbl-aa_loo1) to generate posterior predictions for hypothesis testing. Parameter estimates from the selected model are shown in @tbl-fit_aa1.

To test whether LMA mediated the relationship between growth light intensity and \aax{}, we repeated a similar model comparison procedure but included $\log \mathrm{LMA}$ as an additional fixed effect in the model, as well its interaction with measurement and growth light intensity. The model included a direct effect of growth light intensity on $\log \mathrm{LMA}$ and random phylogeneticlly structured effects of population on $\log \mathrm{LMA}$ and effects of sun treatment on $\log \mathrm{LMA}$. This enabled us to disentangle whether $\log \mathrm{LMA}$ effects \aax{} directly or indirectly because both traits respond to growth light intensity. We integrated over missing LMA data using the `mi()` function in **brms**. Because the evidence strong supported including effects of growth and measurement light intensity on the residual variance in $\mathrm{AA}$, we included these in terms in all models. In total, we fit `r n_model2` models, each with a different combination of fixed and random effect structures. We selected the model with the lowest LOOIC value (@tbl-aa_loo2) to generate posterior predictions for hypothesis testing. Note that LOOIC was calculated only from pointwise likelihood values of the submodel estimating effects of light intensity, light treatments, and population on \aax{}. 
Parameter estimates from the selected model are shown in @tbl-fit_aa2.

@@

The acclimation, plasticity, and constitutive hypotheses make different predictions about the relationship between \aax{} and light intensity, light treatment, and native \ppfd{} among populations. Since these hypotheses are not mutually exclusive, we describe how we assessed support for one, two, or all three hypotheses simultaneously in \autoref{tbl-predictions}. In general, the acclimation hypothesis was supported if \aax{} was greater at high light intensity than low light intensity. The plastic hypothesis was supported if \aax{} was greater in sun leaves than shade leaves. The constitutive hypothesis was supported if population-level \aax{} increased with native \ppfd. In interactive models, we only consider positively reinforcing interactions between high light intensity, sun leaves, and native \ppfd{} because these are the only interactions which could explain why amphistomatous leaves are advantageous in high light habitats.

From the posterior of the selected model (@sec-phylo), we predicted the posterior distribution of the expected \aax{} for each population at each measurement light intensity and growth light intensity. If the 95% confidence intervals for \aax{} did not overlap zero, we considered \aax{} to be significantly different from zero. We also calculated posterior distribution for the average \aax{} among populations in each condition. We tested a linear effect of native light habitat on population-level \aax{} in each treatment condition, integrating over the entire posterior distribution. We calculated the posterior distribution of the slope. If the 95\% confidence interval for the slope did not overlap zero, we considered the slope to be significant.

@@

tribble(
  ~"\\textbf{Hypothesis}", ~"\\textbf{Prediction(s)}", ~"\\textbf{Description}",
  "Acclimation", "$\\mathrm{AA}_{2000} > 0$", "\\aax{} at high light intensity is greater than that at low light intensity",
  "Plasticity", "$\\mathrm{AA}_{\\text{sun}} > 0$", "\\aax{} in sun leaves is greater than that in shade leaves",
  "Constitutive", "$\\beta_{\\mathrm{PAI,AA}} > 0$", "Population-level \\aax{} ($\\mathrm{AA}_\\text{pop}$) decreases with native PAI",
  "Constitutive", "", "$\\mathrm{AA}_\\text{pop} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{pop}}$",
  "Acclimation $\\times$ Plasticity", "$\\mathrm{AA}_{2000,\\text{sun}} > 0$", "\\aax{} is highest at high light intensity in sun leaves",
  "Acclimation $\\times$ Plasticity", "", "",
  "Acclimation $\\times$ Constitutive", "$\\mathrm{AA}_{2000} > 0$", "\\aax{} at high light intensity is greater than that at low light intensity",
  "Acclimation $\\times$ Constitutive", "$\\beta_{\\mathrm{PAI,AA}} > 0$", "Population-level \\aax{} at high light intensity ($\\mathrm{AA}_{\\text{pop},2000}$) decreases with native PAI",
  "Acclimation $\\times$ Constitutive", "", "$\\mathrm{AA}_{\\text{pop},2000} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{pop}} + \\vec{\\beta}_{\\mathrm{AA}, 2000, \\text{pop}}$",
  "Plasticity $\\times$ Constitutive", "$\\mathrm{AA}_{\\text{sun}} > 0$", "\\aax{} in sun leaves is greater than that in shade leaves",
  "Plasticity $\\times$ Constitutive", "$\\beta_{\\mathrm{PAI,AA}} > 0$", "Population-level \\aax{} in sun leaves ($\\mathrm{AA}_{\\text{pop},\\text{sun}}$) decreases with native PAI",
  "Plasticity $\\times$ Constitutive", "", "$\\mathrm{AA}_{\\text{pop},\\text{sun}} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{pop}} + \\vec{\\beta}_{\\mathrm{AA}, \\text{sun}, \\text{pop}}$",
  "Acclimation $\\times$ Plasticity $\\times$ Constitutive", "$\\mathrm{AA}_{2000,\\text{sun}} > 0$", "\\aax{} is highest at high light intensity in sun leaves",
  "Acclimation $\\times$ Plasticity $\\times$ Constitutive", "$\\beta_{\\mathrm{PAI,AA}} > 0$", "Population-level \\aax{} at high light intensity in sun leaves ($\\mathrm{AA}_{\\text{pop},2000,\\text{sun}}$) decreases with native PAI",
  "Acclimation $\\times$ Plasticity $\\times$ Constitutive", "", "$\\mathrm{AA}_{\\text{pop},2000,\\text{sun}} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{pop}} + \\vec{\\beta}_{\\mathrm{AA}, 2000, \\text{pop}} + \\vec{\\beta}_{\\mathrm{AA}, \\text{sun}, \\text{pop}}$"

@@

### Estimating the water cost of hypostomy

Amphistomy advantage implies that hypostomatous leaves would need to lose more water by having higher \gsw{} to achieve the same photosynthetic rate. We estimated this cost of hypostomy approximately. For simpliticity, we focus on a sufficiently small region of the \agcurve{} curve such that we can approximate the relationship between \gsw{} and $A$ as log-log linear. We also assume that slope is same in both amphistomatous and hypostomatous leaves, which was often approximately the case in our experiment (@fig-curves). With these assumptions, the approximate water cost of hypostomy can be derived from the equations:

@@

$$-\log\left(\frac{g_\text{sw,amphi}}{g_\text{sw,hypo}}\right) = \frac{\mathrm{AA}}{\varepsilon_g}.$$
We note that the slope in this case is an elasticity of of $A$ to \gsw{} because both are log-transformed. This quantity will be proportional the water cost of hypostomy assuming that VPD and other environmental variables are held constant and boundary layer conductance are high enough to be negligible.

@@

## Results

@@

```{r fit_aa2, results = 'asis', echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
#| label: tbl-fit_aa2
#| tbl-cap: "Parameter estimates and 95% confidence intervals (CIs) from the posterior distribution of the selected model with leaf mass per area (LMA) as potential mediator variable. Models potentially include fixed, random, and distributional effects of growth light intensity (shade and sun), measurement light intensity (low and high), $\\log \\mathrm{LMA}$, and their interactions on amphistomy advantage ($\\mathrm{AA}$). The distributional effects refer to the effect of factors on residual variation in $\\mathrm{AA}$, denoted $\\sigma$, on a log-link scale. LMA can also be affected by growth light intensity and population. For phylogenetically structured random effects, we report the estimated standard deviation (SD)."

@@

```{r, aa_loo1, echo=FALSE,message=FALSE,warning=FALSE}
#| label: tbl-aa_loo1
#| tbl-cap: 'Leave-one-out cross-validation information criterion (LOOIC) for models of \\aax{} as a function of light intensity, light treatment, and population. Only the top models ($\\Delta \\text{LOOIC} < 2$) are shown. The model with the lowest LOOIC is considered the best fit to the data. The LOOIC column lists the LOOIC value for each model; the $\\Delta$LOOIC column lists the difference between the LOOIC of the corresponding model and the best model (lowest LOOIC). Checkmarks in each row indicate whether the corresponding model included affects of measurement light intensity ("Meas.") or growth light intensity ("Growth") as fixed effects, random effects, or distributional effect on the residual variance. We also test for fixed and random interaction effects ("Inter.") between measurement and growth light intensity. All models included phylogenetically structured random effects of population.'

@@

```{r, aa_loo2, results = "asis", eval=TRUE,echo=FALSE,message=FALSE,warning=FALSE}
#| label: tbl-aa_loo2
#| tbl-cap: 'Leave-one-out cross-validation information criterion (LOOIC) for models of \\aax{} as a function of light intensity, light treatment, leaf mass per area (LMA), and population. Only the top models ($\\Delta~\\text{LOOIC} < 2$) are shown. The model with the lowest LOOIC is considered the best fit to the data. The LOOIC column lists the LOOIC value for each model; the $\\Delta$LOOIC column lists the difference between the LOOIC of the corresponding model and the best model (lowest LOOIC). Checkmarks in each row indicate whether the corresponding model included affects of measurement light intensity ("Meas."), growth light intensity ("Growth"), or LMA as fixed or random effects. We test for fixed interaction effects between measurement and growth light intensity and LMA. We also test for random interaction effects between measurement and growth light intensity. All models included phylogenetically structured random effects of population.'

@@

```{r, curves, echo=FALSE,message=FALSE,warning=FALSE}
#| label: fig-curves
#| fig-cap: "(Figures in separate file) The \\agcurve{} curves, fitted lines, and \\aax{} estimates for every individual plant. The title provides the Tomato Genetics Resource Center accession number, replicate letter, and species names. The subtitle indicates the growth light intensity (sun or shade). Points are raw data, lines are fitted curves, and shaded regions are 95% confidence ribbons for the fitted curves. The Bayesian correlation coefficient $r^2$ is shown to the right of each curve, along with the estimate of \\aax{} $\\pm$ two standard errors for each pair of amphi (orange) and pseudohypo (blue) curves."
ggplot()
```

>>>>

We compared \aax{} among `r n_acc` ecologically diverse populations of wild tomato, including representatives of all described species of *Solanum* sect. *Lycopersicon* and sect. *Lycopersicoides* [@peralta_taxonomy_2008] and the cultivated tomato *S. lycopersicum* var. *lycopersicum* (@tbl-populations). These populations were selected in a previous study because they encompass the breadth of climatic variation in the wild tomato clade [@pease_phylogenomics_2016]. In one case, we substituted a congeric population of the species (*S. galapagense*) because of difficult growing the focal population. We also added an a population of *S. pennellii*. Due to constraints on growth space and time, we spread out measurements over `r n_weeks` weeks from `r start_licor_date` to `r end_licor_date`. Replicates within population were evenly spread out over this period to prevent confounding of temporal variation in growth conditions with variation among populations.

@@

We recorded $\mathrm{PPFD}$ using full spectrum quantum sensors (SQ-500-SS, Apogee Instruments, Logan, Utah, USA); we recorded temperature, RH, and [CO$_2$] using an EE894 sensor (E+E Elektronik, Engerwitzdorf, Austria) protected by a radiation shield. All environmental measurements were taken every 10 minutes from the middle of plant racks at approximately the same height as the leaves we measured. We measured leaf temperature of focal leaves prior to measurement using an infrared radiometer (SI-111-SS, Apogee Instruments, Logan, Utah, USA).

@@

Seeds provided by the Tomato Genetics Resource Center germinated on moist paper in plastic boxes after soaking for 30-60 minutes in a 50% (volume per volume) solution of household bleach and water, followed by a thorough rinse. We transferred seedlings to cell-pack flats containing Pro-Mix BX potting mix (Premier Tech, Rivière-du-Loup, Quebec, Canada) once cotyledons fully emerged, typically within 1-2 weeks of sowing. We grew seeds and seedlings for both sun and shade treatments under the same environmental conditions (12:12 h, `r germ_temp_day`:$\qty{`r germ_temp_night`}{\degreeCelsius}$, `r germ_rh_day`:`r germ_rh_night` RH day:night cycle). LED light provided $\mathrm{PPFD} = \qty{`r germ_ppfd`}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$ (Fluence RAZRx, Austin, Texas, USA) during the germination and seedling stages.

@@

<!-- Not using this currently  -->
<!-- #### Light-saturated photosynthetic rate -->

<!-- In `r percent_aa_and_aq`\% of plants, we measured the maximum photosynthetic rate ($A_\mathrm{max}$) on the same leaflets as \agcurve{} curves. However, when a leaflet was damaged during \agcurve{} curves, we used the next closest leaflet for $A_\mathrm{max}$. Leaves acclimated to high light-intensity (\ppfdequals{2000}), ambient CO$_2$ (\caequals{415}), \rhequals{50}, and \tleafequals{25}. After $A$ and \gsw{} stabilized, we measured $A$ at $\qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$. -->

@@

1. Estimate \agcurve{} curve parameters
2. Estimate \aax{} for both measurement light intensities using \agcurve{} curve parameters
3. Estimate the effects of light intensity, light treatment, and population on \aax{} (acclimation and plasticity hypotheses)
4. Estimate the effects of native light habitat on population-level \aax{} (constitutive hypothesis)

All models were fit using a Bayesian HMC sampling in the probabilistic programming language *Stan* [@stan_development_team_stan_2025] using the *R* package **brms** version `r fit_aa1$version$brms` [@burkner_brms_2017]. We used *CmdStan* version `r fit_aa1$version$cmdstan` and **cmdstanr** version `r fit_aa1$version$cmdstanr` [@gabry_cmdstanr_2025] to interface with *R* version `r getRversion()` [@r_core_team_r:_2025]. We sampled the posterior distribution from a single chain with a minimum of `r aa_args$n_iter_init / 2` iterations after `r aa_args$n_iter_init / 2` warmup iterations. If necessary, we refit models with more iterations to decrease the Gelman-Rubin convergence statistic ($\hat{R}$) [@gelman_inference_1992] to less than `r aa_args[["max_rhat"]]` and the bulk effective sample size (ESS) to greater than `r aa_args[["min_ess"]]` for all model parameters, and there were fewer than `r aa_args[["max_divergent"]]` divergent transitions.

@@

We fit Bayesian mixed effects models with phylogenetically structured random effects to test predictions of competing hypotheses about why amphistomy advantage (\aax) might be greater for leaves in sunny, open habitats. We fit all models combinations of measurement light intensity, growth light intensity, and their interaction as both fixed effects and random effects among populations. In other words, we test whether there were main effects of light intensity treatments and/or whether populations varied in their response to light intensity treatments. In all models, we included population as a phylogenetically structured random effect. We accounted for uncertainty in $\widehat{\mathrm{AA}}$ using the standard deviation of the posterior distribution of each $\widehat{\mathrm{AA}}$ estimate (see above). We crossed fixed and random effect structures with fixed effects of measurement light intensity and/or growth light intensity on the distributional parameter, $\sigma$, the residual variance in \aax{}. We used a robust regression approach by assuming $t$-distributed residuals, which makes estimates less susceptible to extreme values. In total, we fit `r n_model1` models, each with a different combination of fixed, random, and distributional effects. We used the RAxML whole-transcriptome concatenated phylogeny based on 2,745 100-kb genomic windows from @pease_phylogenomics_2016. Two of our populations were not in this tree. We used accession LA1044 in place of LA3909, two populations of *S. galapagense*; LA0750 was added as sister to LA0716, two closely related populations of *S. pennellii*. The node separating LA0750 from LA0716 was placed half-way between the next deepest node. We used the leave-one-out cross-validation information criterion (LOOIC) to compare the fit of models (\autoref{tbl-aa_loo1}) using the *R* package **loo** version `r packageVersion("loo")` [@vehtari_practical_2017] to calculate LOOIC values. We selected the model with the lowest LOOIC value (@tbl-aa_loo1) to generate posterior predictions for hypothesis testing. Parameter estimates from the selected model are shown in @tbl-fit_aa1.

To test whether LMA mediated the relationship between growth light intensity and \aax{}, we repeated a similar model comparison procedure but included $\log \mathrm{LMA}$ as an additional fixed effect in the model, as well its interaction with measurement and growth light intensity. The model included a direct effect of growth light intensity on $\log \mathrm{LMA}$ and random phylogeneticlly structured effects of population on $\log \mathrm{LMA}$ and effects of sun treatment on $\log \mathrm{LMA}$. This enabled us to disentangle whether $\log \mathrm{LMA}$ effects \aax{} directly or indirectly because both traits respond to growth light intensity. We integrated over missing LMA data using the `mi()` function in **brms**. Because the evidence from the first set of models strongly supported including effects of growth and measurement light intensity on the residual variance in $\mathrm{AA}$, we included these in terms in all models. In total, we fit `r n_model2` models, each with a different combination of fixed and random effect structures. We selected the model with the lowest LOOIC value (@tbl-aa_loo2) to generate posterior predictions for hypothesis testing. Note that LOOIC was calculated only from pointwise likelihood values of the submodel estimating effects of light intensity, light treatments, and population on \aax{}. 
Parameter estimates from the selected model are shown in @tbl-fit_aa2.

@@

The acclimation, plasticity, and constitutive hypotheses make different predictions about the relationship between \aax{} and light intensity, light treatment, and native $\mathrm{PAI}$ among populations. Since these hypotheses are not mutually exclusive, we describe how we assessed support for one, two, or all three hypotheses simultaneously in \autoref{tbl-predictions}. In general, the acclimation hypothesis was supported if \aax{} was greater at high light intensity than low light intensity. The plastic hypothesis was supported if \aax{} was greater in sun leaves than shade leaves. The constitutive hypothesis was supported if population-level \aax{} increased with native \ppfd. In interactive models, we only consider positively reinforcing interactions between high light intensity, sun leaves, and native \ppfd{} because these are the only interactions which could explain why amphistomatous leaves are advantageous in high light habitats.

From the model selected using LOOIC, we predicted the posterior distribution of the expected \aax{} for each population at each measurement light intensity and growth light intensity. If the 95% confidence intervals for \aax{} did not overlap zero, we considered \aax{} to be significantly different from zero. We also calculated posterior distribution for the average \aax{} among populations in each condition. We tested a linear effect of native light habitat on population-level \aax{} in each treatment condition, integrating over the entire posterior distribution to calculate a posterior distribution of the slope. If the 95\% confidence interval for the slope did not overlap zero, we considered the slope to be significant.

@@

tribble(
  ~"\\textbf{Hypothesis}", ~"\\textbf{Prediction(s)}", ~"\\textbf{Description}",
  "Acclimation", "$\\mathrm{AA}_{2000} > 0$", "\\aax{} at high light intensity is greater than that at low light intensity",
  "Plasticity", "$\\mathrm{AA}_{\\text{sun}} > 0$", "\\aax{} in sun leaves is greater than that in shade leaves",
  "Constitutive", "$\\beta_{\\mathrm{PAI,AA}} < 0$", "Population-level \\aax{} ($\\mathrm{AA}_\\text{pop}$) decreases with native PAI",
  "Constitutive", "", "$\\mathrm{AA}_\\text{pop} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{pop}}$",
  "Acclimation $\\times$ Plasticity", "$\\mathrm{AA}_{2000,\\text{sun}} > 0$", "\\aax{} is highest at high light intensity in sun leaves",
  "Acclimation $\\times$ Plasticity", "", "",
  "Acclimation $\\times$ Constitutive", "$\\mathrm{AA}_{2000} > 0$", "\\aax{} at high light intensity is greater than that at low light intensity",
  "Acclimation $\\times$ Constitutive", "$\\beta_{\\mathrm{PAI,AA}} < 0$", "Population-level \\aax{} at high light intensity ($\\mathrm{AA}_{\\text{pop},2000}$) decreases with native PAI",
  "Acclimation $\\times$ Constitutive", "", "$\\mathrm{AA}_{\\text{pop},2000} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{pop}} + \\vec{\\beta}_{\\mathrm{AA}, 2000, \\text{pop}}$",
  "Plasticity $\\times$ Constitutive", "$\\mathrm{AA}_{\\text{sun}} > 0$", "\\aax{} in sun leaves is greater than that in shade leaves",
  "Plasticity $\\times$ Constitutive", "$\\beta_{\\mathrm{PAI,AA}} < 0$", "Population-level \\aax{} in sun leaves ($\\mathrm{AA}_{\\text{pop},\\text{sun}}$) decreases with native PAI",
  "Plasticity $\\times$ Constitutive", "", "$\\mathrm{AA}_{\\text{pop},\\text{sun}} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{pop}} + \\vec{\\beta}_{\\mathrm{AA}, \\text{sun}, \\text{pop}}$",
  "Acclimation $\\times$ Plasticity $\\times$ Constitutive", "$\\mathrm{AA}_{2000,\\text{sun}} > 0$", "\\aax{} is highest at high light intensity in sun leaves",
  "Acclimation $\\times$ Plasticity $\\times$ Constitutive", "$\\beta_{\\mathrm{PAI,AA}} > 0$", "Population-level \\aax{} at high light intensity in sun leaves ($\\mathrm{AA}_{\\text{pop},2000,\\text{sun}}$) decreases with native PAI",
  "Acclimation $\\times$ Plasticity $\\times$ Constitutive", "", "$\\mathrm{AA}_{\\text{pop},2000,\\text{sun}} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{pop}} + \\vec{\\beta}_{\\mathrm{AA}, 2000, \\text{pop}} + \\vec{\\beta}_{\\mathrm{AA}, \\text{sun}, \\text{pop}}$"

@@

### The water cost of hypostomy

Amphistomy advantage implies that hypostomatous leaves would need to lose more water by having higher \gsw{} to achieve the same photosynthetic rate as an otherwise identical amphistomatous leaf. We estimated this cost of hypostomy approximately. For simplicity, we focus on a sufficiently small region of the \agcurve{} curve such that we can approximate the relationship between \gsw{} and $A$ as log-log linear. We also assume that the slope is the same in both amphistomatous and hypostomatous leaves, which was often approximately the case in our experiment (@fig-curves). With these assumptions, the approximate water cost of hypostomy can be derived from the equations:

@@

$$-\log\left(\frac{g_\text{sw,amphi}}{g_\text{sw,hypo}}\right) = \frac{\mathrm{AA}}{\varepsilon_g}.$$
We note that the slope in this case is the elasticity of $A$ to \gsw{} because both are log-transformed. This quantity will be proportional the water cost of hypostomy assuming that VPD and other environmental variables are held constant and the boundary layer resistance is negligible.

@@

@@

```{r fit_aa2, results = 'asis', echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
#| label: tbl-fit_aa2
#| tbl-cap: "Parameter estimates and 95% confidence intervals (CIs) from the posterior distribution of the selected model with leaf mass per area (LMA) as potential mediator variable. Models potentially include fixed and random effects of growth light intensity (shade and sun), measurement light intensity (low and high), $\\log \\mathrm{LMA}$, and their interactions on amphistomy advantage ($\\mathrm{AA}$). LMA can also be affected by growth light intensity and population. For phylogenetically structured random effects, we report the estimated standard deviation (SD)."

@@

```{r, aa_loo1, echo=FALSE,message=FALSE,warning=FALSE}
#| label: tbl-aa_loo1
#| tbl-cap: 'Leave-one-out cross-validation information criterion (LOOIC) for models of \\aax{} as a function of light intensity, light treatment, and population. Only the top models ($\\Delta \\text{LOOIC} < 2$) are shown. The model with the lowest LOOIC was selected for generating posterior predictions to test hypotheses. The LOOIC column lists the LOOIC value for each model; the $\\Delta$LOOIC column lists the difference between the LOOIC of the corresponding model and the best model (lowest LOOIC). Checkmarks in each row indicate whether the corresponding model included affects of measurement light intensity ("Meas.") or growth light intensity ("Growth") as fixed effects, random effects, or distributional effect on the residual variance. We also test for fixed and random interaction effects ("Inter.") between measurement and growth light intensity. All models included phylogenetically structured random effects of population.'

@@

```{r, aa_loo2, results = "asis", eval=TRUE,echo=FALSE,message=FALSE,warning=FALSE}
#| label: tbl-aa_loo2
#| tbl-cap: 'Leave-one-out cross-validation information criterion (LOOIC) for models of \\aax{} as a function of light intensity, light treatment, leaf mass per area (LMA), and population. Only the top models ($\\Delta~\\text{LOOIC} < 2$) are shown. The model with the lowest LOOIC was selected for generating posterior predictions to test hypotheses. The LOOIC column lists the LOOIC value for each model; the $\\Delta$LOOIC column lists the difference between the LOOIC of the corresponding model and the best model (lowest LOOIC). Checkmarks in each row indicate whether the corresponding model included affects of measurement light intensity ("Meas."), growth light intensity ("Growth"), or LMA as fixed or random effects. We test for fixed interaction effects between measurement and growth light intensity and LMA. We also test for random interaction effects between measurement and growth light intensity. All models included phylogenetically structured random effects of population.'

@@

```{r, curves, echo=FALSE,message=FALSE,warning=FALSE}
#| label: fig-curves
#| fig-cap: "(Figures in separate file) The \\agcurve{} curves, fitted lines, and \\aax{} estimates for every individual plant. The title provides the Tomato Genetics Resource Center accession number, replicate letter, and species names. The subtitle indicates the growth light intensity (sun or shade). Points are raw data, lines are fitted curves, and shaded regions are 95% confidence ribbons for the fitted curves. The Bayesian correlation coefficient, $r^2$, is shown to the right of each curve, along with the estimate of \\aax{} $\\pm$ two standard errors for each pair of amphi (orange) and pseudohypo (blue) curves."
ggplot()
```
